Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-11 manu-draft.Rmd
adding co-authors
edits to benthic monitoring

---
title: "Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.  For benthic communities, sediments were sampled in April during the discharge near Port Manatee and surrounding waters.  These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results suggested a mix of conditions dominated by "fair" and "good" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].  Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).  Additional sampling is planned later at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.  Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.

>>>>

---
title: "Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.  For benthic communities, sediments were sampled in April during the discharge near Port Manatee and surrounding waters.  These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "healthy", "intermediate", and "degraded" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].  Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).  Additional sampling occurred in September at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.  Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
